<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Reproducing an online report using ImmuneSpaceR: Correlation between HAI and flow cytometry in SDY269}
-->
# Correlating HAI with flow cytometry and ELISPOT results in SDY269

ImmuneSpaceR code produces consistent results, regardless of whether it is being executed from a module or a UI based report on the server or on a local machine. This vignette reproduces a report available on <a href="https://www.immunespace.org/reports/Studies/SDY269/runReport.view?reportId=module%3ASDY269%2Freports%2Fschemas%2Fhai_flow_elispot.Rmd&redirectUrl=%2Fproject%2FStudies%2FSDY269%2Fbegin.view%3FpageId%3DReports">www.immunespace.org</a> using the same code.

### Summary
This report investigate the association between the number influenza-specific cells measured by ELISPOT measured at day 7 with the number of plasmablast measured by flow cytometry and day 7 and the HAI response measured at day 28 (log-fold day28/day0). It basically reproduces Figure 1 d-e) of <a href="http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3140559/" target="_BLANK"> Nakaya et al. (2011)</a> published as part of the original study.


```{r knitr-opts, echo = FALSE, message = FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(cache=FALSE, echo=TRUE, message=FALSE, warning=FALSE,
               fig.width=8, fig.height=4, dpi=100, fig.align="center")
```

#### Load ImmuneSpaceR and other libraries
```{r libraries, cache=FALSE}
library(ImmuneSpaceR)
library(ggplot2)
library(data.table)
library(ggthemr)
ggthemr('solarized')
```


#### Connect to the study and get datasets
```{r connection}
study <- CreateConnection(c("SDY269"))
dt_hai <- study$getDataset("hai", reload=TRUE)
dt_fcs <- study$getDataset("fcs_analyzed_result", reload=TRUE)
dt_elispot <- study$getDataset("elispot", reload=TRUE)
```
#### Transform data
```{r data-subset}
# Compute max fold change for HAI, and remove time zero
dt_hai <- dt_hai[,hai_response:=value_reported/value_reported[study_time_collected==0],
                 by="virus_strain,arm_name,subject_accession"][study_time_collected==28]
dt_hai <- dt_hai[,list(hai_response=max(hai_response)),by="arm_name,subject_accession"]

# Define variable for ELISPOT, keep only the IgG class
dt_elispot <- dt_elispot[,elispot_response:=spot_number_reported+1][study_time_collected==7 & analyte=="IgG"]
# Compute % plasmablasts
dt_fcs <- dt_fcs[,fcs_response:=(as.double(population_cell_number)+1) /
                   as.double(base_parent_population)][study_time_collected==7]
```

#### Merge data and phenodata
```{r merging}
# Let's key the different datasets
setkeyv(dt_hai, c("subject_accession"))
setkeyv(dt_fcs, c("subject_accession"))
setkeyv(dt_elispot, c("subject_accession"))
dt_all <- dt_hai[dt_fcs, nomatch=0][dt_elispot, nomatch=0]
```


The figure below shows the absolute number of plasmablast cells measured by flow cytometry vs. the number of frequency of influenza-specific cells measured by ELISPOT. 
```{r plot1, dev='CairoPNG'}
ggplot(dt_all, aes(x=as.double(fcs_response), y=elispot_response, color=arm_name)) +
  geom_point() + scale_y_log10() + scale_x_log10() + geom_smooth(method="lm") +
  xlab("Total plasmablasts (%)") + ylab("Influenza specific cells\n (per 10^6 PBMCs)")
```

The figure below shows the HAI fold increase over baseline vs. the number of frequency of influenza-specific cells measured by ELISPOT. 
```{r plot2, dev='CairoPNG'}
ggplot(dt_all, aes(x=as.double(hai_response), y=elispot_response, color=arm_name)) +
  geom_point() + scale_x_continuous(trans="log2") + scale_y_log10() +
  geom_smooth(method="lm") + xlab("HAI fold") + ylab("Influenza specific cells\n (per 10^6 PBMCs)")
```


In each case, we observe good correlations between the different responses, at least for the TIV cohort.
